### 2020-03-11 ropensci/rclean_technote.Rmd
Creates rclean_technote.Rmd

---
title: 'Rclean: A Tool for Writing Cleaner, More Transparent Code'
tags:
  - R
  - reproducibility
  - transparency
  - code cleaning
  - data provenance
authors:
  - name: Matthew K. Lau
    orcid: 0000-0003-3758-2406
    affiliation: 1
  - name: Thomas F. J.-M. Pasquier
    orcid: 0000-0001-6876-1306
    affiliation: "2, 3" 
  - name: Margo Seltzer
    orcid: 0000-0002-2165-4658
    affiliation: "4"
affiliations:
 - name: Harvard Forest, Harvard University 
   index: 1
 - name: Department of Computer Science, University of Bristol 
   index: 2
 - name: School of Engineering and Applied Science, Harvard University
   index: 3
 - name: Department of Computer Science, University of British Columbia
   index: 4
date: 
bibliography: paper.bib
---

# Introduction

The growth of programming in the sciences has been explosive in the
last decade. This has facilitated the rapid advancement of science
through the agile development of computational tools. However,
concerns have begun to surface about the reproducibility of scientific
research in general [@Peng2011 @Baker2016] and the potential issues
stemming from issues with analytical software [@Pasquier2017
@Stodden2018]. Specifically, there is a growing recognition across
disciplines that simply making data and software "available" is not
enough and that there is a need to improve the transparency and
stability of scientific software [@Pasquier2018].

At the core of the growth of scientific computation, the ``R``
statistical programming language has grown exponentially to become one
of the top ten programming languages in use today. At it's root R is a
*statistical* programming language. That is, it was designed for use
in analytical workflows; and the majority of the R community is
focused on producing code for idiosyncratic projects that are
*results* oriented. Also, R's design is intentionally at a level that
abstracts many aspects of programming that would otherwise act as a
barrier to entry for many users. This is good in that there are many
people who use R to their benefit with little to no formal training in
computer science or software engineering, but these same users can
also be frequently frustrated by code that is fragile, buggy and
complicated enough to quickly become obtuse even to the authors. The
stability, reproducibility and re-use of scientific analyses in R
would be improved by refactoring, which is a common practice in
software engineering [@Martin2009CleanCraftsmanship]. From this
perspective, tools that can lower the time and energy required to
refactor analytical scripts and otherwise help to "clean" code, but
abstracted enough to be easily accessible, could have a significant
impact on scientific reproducibility across all disciplines
[@Visser2015].

To provide support for easier refactoring in R, we have created
``Rclean``. The ``Rclean`` package provides tools to automatically
reduce a script to the parts that are specifically relevant to a
research product (e.g.  a scientific report, academic talk, research
article, etc.). Although potentially useful to all R coders, it was
designed to ease refactoring for scientists that use R but do not have
formal training in software engineering. Here, we detail the structure
of the package's API, describe the general workflow illustrated by an
example use case and provide some background on how data provenance
enables the underlying functionality of the package. We then end with
a discussion of future applications of data provenance in the context
of "code cleaning" and the potential integration with other software
engineering tools for the R community.

# Methods

The goal of `Rclean` is to provide a set of tools that help someone
reduce and organize code based on results. More often then not, when
someone is writing an R script, the intent is to produce a set of
results, such as a statistical analysis, figure, table, etc. This set
of results is always a subset of a much larger set of possible ways to
explore a dataset, as there are many statistical approaches and tests,
let alone ways to create visualizations and other representations of
patterns in data. This commonly leads to lengthy, complicated scripts
from which researchers manually subset results, but likely never to be
refactored because of the difficulty in disentangling the code needed
to produce some results and not others. The 'Rclean' package uses an
automated technique based on data provenance (details below) to
analyze existing scripts and provide ways to identify and extract code
to produce a desired output.

## The API

To keep the process simple and
straight-forward much of this process has been abstracted for the
user, and the API has been kept to a minimum set of functions that
enable a user to conduct the following basic workflow:

1. Obtain the "cleaned" code for a result(s).
2. Transfer the code to a new context (e.g. a new script,
function, reproducible example, web-app, etc.).
3. Get information about the possible results and repeat as needed.

The package's main functions are `clean` and `keep`. When provided a
file path to a script and the name of a result (or a set of results),
`clean` analyzes the script's code and extracts the lines of code
required to produce the results. By default, code is formatted
following general best practices recommended by the Tidyverse via the
`styler` package [@R-styler]. This code can then be passed to the
`keep` function, which can either write the code to disk or copy the
code to the user's clipboard (if no output file path is supplied) and
the user can paste the code into another location (e.g. a script
editor).

In the process of cleaning a script, it is likely that a user will not
know the precise objects they want and might require some help
analyzing it. There are several functions to help with this
process. The `get_vars` function will return a list of possible
results for a given script at a supplied file path. This is obviously
an important step, and justifiably, the default behavior of the
`clean` function is to run `get_vars` if no results are supplied. To
help with limiting and checking the selection of results, the
`code_graph` function creates a network graph of the relationships
among the various results and lines of code in the script, which is
the function used to create the figure demonstrating data provenance
\@ref(fig:prov-graph). Last, the `get_libs` function can be used to detect
the packages that a given script depends on, which it will return as
coded library calls that can be inserted into a cleaned script.

## Data Provenance

All of these processes rely on the generation of data provenance.  The
term provenance means information about the origins of some
object. Data provenance is a formal representation of the execution of
a computational process (https://www.w3.org/TR/prov-dm/), to
rigorously determine the the unique computational pathway from inputs
to results [@Carata2014]. To avoid confusion, note that "data" in this
context is used in a broad sense to include all of the information
generated during computation, not just the data that are collected in
a research project that are used as input to an analysis. Having the
formalized, mathematically rigorous representation that data
provenance provides guarantees that the analyses that `Rclean`
conducts are theoretically sound. Most importantly, because the
relationships defined by the provenance can be represented as a graph,
it is possible to apply network search algorithms to determine the
minimum and sufficient code needed to generate the chosen result in
the `clean` function.

@@

Relatedly, it is important to point out that `Rclean` *does not* keep
comments present in code. This could be seen as a limitation of the
data provenance, which currently does not assign them a
relationships. Therefore, although there is often very useful or even
invaluable information in comments, the `clean` function removes
them. This is a general issue with automated methods for detecting the
relationships between comments and the code itself. Comments at the
end of lines are typically relevant to the line they are on, but this
is not a requirement and could refer to other lines. Also, comments
occupying their own lines usually refer to the following lines, but
this is also not necessarily the case. As `clean` depends on the
unambiguous determination of relationships in the production of
results, it cannot operate automatically on comments. However,
comments in the original code remain untouched and can be used to
inform the reduced code. Also, as the `clean` function is oriented
toward isolating code based on a specific result, the resulting code
tends to naturally support the generation of new comments that are
higher level (e.g. "The following produces a plot of the mean response
of each treatment group."), and lower level comments are not necessary
because the code is simpler and clearer.

In the future, it would also be useful to extend the existing
framework to support other provenance methods. One such possibility is
*retrospective provenance*, which tracks a computational process as it
is executing. Through this active, concurrent monitoring,
retrospective provenance can gather information that static
prospective provenance can't. Greater details of the computational
process would enable other features that could address some
challenges, such as processing information from comments, parsing
control statements and replicating random processes. However, using
retrospective provenance comes at a cost. In order to gather it, the
script needs to be executed. When scripts are computationally
intensive or contain bugs that stop execution, then retrospective
provenance can not be obtained for part or all of the code. Some work
has already been done in the direction of implementing retrospective
provenance for code cleaning in R (see http://end-to-end-provenance.github.io).

# Results

## Example

Data analysis can be messy and complicated, so it's no wonder that the
code often reflects this. "What did I measure? What analyses are
relevant to them? Do I need to transform the data? What's the function
for the analysis I want to run?" This is why having a way to isolate
code based on variables can be valuable. The following is an example
of a script that has some complications. As you can see, although the
script is not extremely long or complicated, it's difficult enough to
make it frustrating to visualize it in its entirety and pick through
it.

@@

# Discussion

We have created ``Rclean`` to provide a simple, easy to use tool for
scientists who would like help refactoring code. Using ``Rclean`` the
code necessary to produce a specified result (e.g., an object stored
in memory or a table or figure written to disk) can be easily and
*reliably* isolated even when tangled with code for other
results. This functionality is enabled by graph analysis of data
provenance collected from the user's script. This is likely to be a
broadly useful tool as statistical programming becomes more common
across the sciences. Tools, such as this, that make it easier to
produce transparent, accessible code will be an important aid for
improving scientific reproducibility.

Although the workflow described is impactful, we see promise in
extending it to interface with other clean code and reproducibility
tools. One example is the `reprex` package, which provides a simple
API for sharing reproducible examples [@R-reprex]. Another possibility
is to help transition scripts to function, package and workflow
creation and refactoring via toolboxes like `drake`
[@R-drake]. `Rclean` could provide a reliable way to extract parts of
a larger script that would be piped to a simplified reproducible
example, in the case of `reprex`, or, since it can isolate the code
from inputs to one or more outputs, be used to extract all of the
components needed to write one or more functions that would be a part
of a package or workflow, as is the goal of `drake`.

To conclude, we hope that `Rclean` makes writing scientific software
easier for the R community. We look forward to feedback and help with
extending its application, particularly in the area of
reproducibility. To get involved, report bugs, suggest features,
please visit the project page.

# Acknowledgments

This work was improved by discussions with ecologists at Harvard
Forest and through the helpful review provided by the ROpenSci
community, particularly Anna Krystalli, Will Landau and Clemens
Schmid. Much of the work was funded by US National Science Foundation
grant SSI-1450277 for applications of End-to-End Data Provenance.

>>>>

Rclean: Automated Tools for Cleaning R Code
===========================================

> Leave the code cleaner than you found it. 
-- R.C. Martin in *Clean Code*

Data analysis can be messy and complicated, so it's no wonder that the
code often reflects this. "What did I measure? What analyses are
relevant to them? Do I need to transform the data? What's the function
for the analysis I want to run?" In addition, although many
researchers see the value in learning to write software, the entry
level is still exceedingly high. The `R` language has become very
popular among scientists and analysts because of the lower threshold,
which is good in that it empowers science. The downside is that
learning to create good software is a skill that takes a considerable
amount of time to learn, which means that many scientists write code
that will get the job "done" but lacks transparency and is highly
unstable (i.e. buggy).

The goal of `Rclean` is to provide a tool that help someone reduce and
organize code based on results.  More often then not, when someone is
writing an R script, the intent is to produce a set of results, such
as a statistical analysis, figure or table. This set of results is
nearly always a subset of a much larger set of possible ways to
explore a dataset, as there are many statistical approaches and tests,
let alone ways to create visualizations and other representations of
patterns in data. This commonly leads to lengthy, complicated scripts
from which researchers manually subset results, but likely never to be
refactored because of the difficulty in disentangling the code needed
to produce some results and not others.

The `Rclean` package uses an automated technique based on data
provenance and network algorithms to analyze existing scripts and
provide ways to identify and extract code to produce a desired output
without having to manually conduct the search through the code line by
line. `Rclean` was designed to ease refactoring for scientists that
use R but do not have formal training in software engineering and
specifically with the "art" of creating clean code; however, because
manually culling code is tedious and potentially leads to errors,
`Rclean` can be a useful tool for everyone at all levels of coding
ability.

# The Provenance Engine Under the Hood

The process that `Rclean` uses relies on the generation of data
provenance.  The term provenance means information about the origins
of some object. Data provenance is a formal representation of the
execution of a computational process (https://www.w3.org/TR/prov-dm/),
to rigorously determine the the unique computational pathway from
inputs to results [@Carata2014]. To avoid confusion, note that "data"
in this context is used in a broad sense to include all of the
information generated during computation, not just the data that are
collected in a research project that are used as input to an
analysis. Having the formalized, mathematically rigorous
representation that data provenance provides guarantees that the
analyses that `Rclean` conducts are theoretically sound. Most
importantly, because the relationships defined by the provenance can
be represented as a graph, it is possible to apply network search
algorithms to determine the minimum and sufficient code needed to
generate the chosen result in the `clean` function.

@@

As you can see, the provenance provides a network representation of
relationships among functions and objects. This is very powerful
because we can now apply algorithms analyze the R script with respect
to our results. The main function of `Rclean`, called `clean`, takes
the provenance and applies a depth first network search algorithm to
determine the pathways leading from inputs to outputs, excluding any
objects or lines of code that do not fall along that pathway and are,
therefore, not necessary to produce the desired set of results. As
demonstrated in the next section, through this process, we can produce
a new script with the minimal code necessary to generate the output we
want.

# Using `Rclean` to clean your code

The following is an example of a script that has some "complications."
As you can see, although the script is not extremely long or
complicated, it's difficult enough to make it frustrating to visualize
it in its entirety and pick through it.

@@

# Concluding Remarks and Future Work

We have created ``Rclean`` to provide a simple, easy to use tool for
scientists who would like help refactoring code. Using ``Rclean`` the
code necessary to produce a specified result (e.g., an object stored
in memory or a table or figure written to disk) can be easily and
*reliably* isolated even when tangled with code for other
results. Tools, such as this, that make it easier to produce
transparent, accessible code will be an important aid for improving
scientific reproducibility.

The `clean` function provides an effective way to remove code that is
unwanted; however, many researchers are wary about doing this exact
thing for at least a few reasons. Perhaps the top reason is that the
main goal of an analysis is the results and taking time to craft
transparent, dependable software is not the priority. As such, taking
time to go back through a script and remove code is time
wasted. Relatedly, for most researchers the best way to keep track of
the various analyses that they have explored is to keep them in the
script, as they do not use a rigorous version control system but
instead rely on file backups and informal versioning. Although we
can't give researchers more hours in the day, providing an easier and
more reliable means to remove unused code will lower the barrier to
creating better, cleaner code. Combined with the increasing use of
version control systems and digital notebooks, the practice of
"saving" analytical ideas in a script will become less common and code
quality will increase.

Relatedly, it is important to point out that `Rclean` *does not* keep
comments present in code. This is a limitation of the data provenance,
which currently does not assign them a relationships. Therefore,
although there is often very useful or even invaluable information in
comments, the `clean` function removes them. This is a general issue
with automated methods for detecting the relationships between
comments and the code itself. Comments at the end of lines are
typically relevant to the line they are on, but this is not a
requirement and could refer to other lines. Also, comments occupying
their own lines usually refer to the following lines, but this is also
not necessarily the case. As `clean` depends on the unambiguous
determination of relationships in the production of results, it cannot
operate automatically on comments. However, comments in the original
code remain untouched and can be used to inform the reduced
code. Also, as the `clean` function is oriented toward isolating code
based on a specific result, the resulting code tends to naturally
support the generation of new comments that are higher level
(e.g. "The following produces a plot of the mean response of each
treatment group."), and lower level comments are not necessary because
the code is simpler and clearer. Lastly, as stated above, comments can
serve an important role in coding, it is worth reflecting on the
statement in R.C. Martin's book `Clean Code: A Handbook of Agile
Software Craftsmanship` that "Comments do not compensate for bad
code."

In the future, it would also be useful to extend the existing
framework to support other provenance methods. One possibility is
*retrospective provenance*, which tracks a computational process as it
is executing. Through this active, concurrent monitoring,
retrospective provenance can gather information that static
prospective provenance can't. Greater details of the computational
process would enable other features that could address some
challenges, such as processing information from comments, parsing
control statements and replicating random processes. However, using
retrospective provenance comes at a cost. In order to gather it, the
script needs to be executed. When scripts are computationally
intensive or contain bugs that stop execution, then retrospective
provenance can not be obtained for part or all of the code. Some work
has already been done in the direction of implementing retrospective
provenance for code cleaning in R (see
http://end-to-end-provenance.github.io).

Although the current implementation of Rclean for minimizing code is
useful on its own, we see promise in connecting it with other clean
code and reproducibility tools. One example is the `reprex` package,
which provides a simple API for sharing reproducible examples
[@R-reprex]. Another possibility is to help transition scripts to
function, package and workflow creation and refactoring via toolboxes
like `drake` [@R-drake]. `Rclean` could provide a reliable way to
extract parts of a larger script that would be piped to a simplified
reproducible example, in the case of `reprex`, or, since it can
isolate the code from inputs to one or more outputs, be used to
extract all of the components needed to write one or more functions
that would be a part of a package or workflow, as is the goal of
`drake`.

To conclude, we hope that `Rclean` makes writing scientific software
easier for the R community. As an open-source project, we look forward
to feedback and help with extending it, particularly in the area of
reproducibility. We invite people to use the package and get involved
by reporting bugs and suggesting or contributing features. For more
information please visit the project page on Github.
